---
title: "p8105_hw2_kpp2126"
author: "Kevin P. Patterson"
date: "2022-09-26"
output: github_document
---

```{r setup}
library(tidyverse)
```

## Problem 1

*Read and clean the data; 
*retain line, station_name, station latitude / longitude, routes served (route1:route11), entry, vending, entrance_type, and ADA compliance (ada). 
*Convert the entry variable from character (YES vs NO) to a logical variable (the ifelse or recode function may be useful).
```{r clean data}
#getwd()
transit_df =
  read.csv(
    "../../datascience/p8105_hw2_kpp2126/NYC_Transit_Subway_Entrance_And_Exit_Data.csv") %>%
  janitor::clean_names() %>%
  select(
    line, station_name, station_latitude, station_longitude, route1:route11, entry, vending, entrance_type, ada) %>%
  mutate(
    entry = ifelse(entry == "YES", TRUE, FALSE))
```

*Write a short paragraph about this dataset – explain briefly what variables the dataset contains, describe your data cleaning steps so far, and give the dimension (rows x columns) of the resulting dataset. 
*Are these data tidy?

*This dataset has `dim(trans_df)` observations and column variables respectively.

Answer the following questions using these data:

*How many distinct stations are there? Note that stations are identified both by name and by line (e.g. 125th St 8th Avenue; 125st Broadway; 125st Lenox); the distinct function may be useful here.




