---
title: "p8105_hw2_kpp2126"
author: "Kevin P. Patterson"
date: "2022-09-26"
output: github_document
---

```{r setup}
library(tidyverse)
library(dplyr)
library(readr)
library(readxl)
```

## Problem 1

*Read and clean the data; 
(1)retain line, station_name, station latitude / longitude, routes served (route1:route11), entry, vending, entrance_type, and ADA compliance (ada). 
(2)Convert the entry variable from character (YES vs NO) to a logical variable (the `ifelse` or `recode` function may be useful).
```{r clean data}
transit_df =
  read.csv(
    "../../datascience/p8105_hw2_kpp2126/NYC_Transit_Subway_Entrance_And_Exit_Data.csv") %>%
  janitor::clean_names() %>%
  select(line, station_name, station_latitude, station_longitude, route1:route11, entry, vending, entrance_type, ada) %>%
  mutate(entry = ifelse(entry == "YES", TRUE, FALSE))

#Tidying transit data frame
transit_df_tidy = transit_df %>%
  mutate_at(c('route8', 'route9', 'route10', 'route11'), as.character)

#Need to pivot longer to make route its own variable
transit_df_tidy = 
  pivot_longer(
    transit_df_tidy,
    route1:route11,
    names_to = "route",
    values_to = "train")
```

*Write a short paragraph about this dataset – explain briefly what variables the dataset contains, describe your data cleaning steps so far, and give the dimension (rows x columns) of the resulting dataset. 
*Are these data tidy?

The resulting dataset has `r dim(transit_df)` observations and column variables respectively. Among the `r ncol(transit_df)` columns,  there are a mixture of numeric, logical, and character vectors in the dataframe. Despite there being `r nrow(transit_df)` observations, these do not represent the true distinct stations because there are duplicates. This is due to the dataset being in wide format, made to accommodate for columns that were in the larger dataset. Thus, the current dataframe is not tidy and I run an example of tidying the dataframe to a longer format, making the `route1:route11` columns into a single variable and its values to another variable called `train`. The dimensions of the tidied dataframe, `transit_df_tidy` are `r dim(transit_df_tidy)` observations and column variables respectively, reflecting the longer formatted dataframe.

*How many distinct stations are there? Note that stations are identified both by name and by line (e.g. 125th St 8th Avenue; 125st Broadway; 125st Lenox); the `distinct` function may be useful here.


```{r distinct stations}
distinct_stations <- transit_df %>%
  select(station_name, line) %>%
  distinct
```
There are `r pull(count(distinct_stations))` distinct stations that are identified both by name and by line.

*How many stations are ADA compliant?
```{r ada compliance}
ADA_stations <- transit_df %>%
  filter(ada == TRUE) %>%
  select(station_name, line) %>%
  distinct
```
There are `r pull(count(ADA_stations))` distinct stations, identified both by name and by line, that are ADA compliant.

*What proportion of station entrances / exits without vending allow entrance?
```{r proportion}
transit_df %>%
  filter(vending == "NO") %>%
  pull(entry) %>%
  mean
```
The proportion of station entrances/exits without vending that allow entrance is `r mean(transit_df[transit_df$vending == 'NO', 'entry'])`.

(1)Reformat data so that route number and route name are distinct variables. 
*How many distinct stations serve the A train? Of the stations that serve the A train, how many are ADA compliant?
```{r reformat}
distinct_stations_A <- transit_df_tidy %>% #assigned to a df to call in inlineR
  filter(train == "A") %>%
  select(station_name, line) %>%
  distinct

distinct_stations_A_ADA <- transit_df_tidy %>% #assigned to a df to call in inlineR
  filter(train == "A", ada == TRUE) %>%
  select(station_name, line) %>%
  distinct
```
There are `r count(distinct_stations_A)` distinct stations that serve the A train, and of these stations there are `r count(distinct_stations_A_ADA)` stations are ADA compliant.

## Problem 2

*Read and clean the Mr. Trash Wheel sheet:

(1)specify the `sheet = "Mr Trash Wheel` in the Excel file and to omit non-data entries (rows with notes / figures; columns containing notes) using `skip` and `range` in read_excel
(2)use reasonable variable names
(3)omit rows that do not include dumpster-specific data with `drop_na(dumpster)`
(4)round the number of sports balls to the nearest integer and converts the result to an integer variable (using as.integer)
(4a)To keep track of which Trash Wheel is which, you may need to add an additional variable to both datasets before combining.

```{r load clean Mr. Trash Wheel}
trashwheel_df =
  readxl::read_excel(
    "../../datascience/p8105_hw2_kpp2126/Trash Wheel Collection Data.xlsx",
    sheet = "Mr. Trash Wheel",
    skip = 1, #may be extra, but added as cautionary measure
    range = cellranger::cell_cols("A:N")) %>%
  janitor::clean_names() %>%
  drop_na(dumpster) %>% #specified for the column opposed to dropping the row for any NA's across
  mutate(sports_balls = round(sports_balls),
         sports_balls = as.integer(sports_balls),
         homes_powered = as.integer(homes_powered)) %>% #matched this with professortrashwheel_df for binding as cautionary measure
  add_column(trash_wheel_type = "mister") #identifying the dataframe before binding
```


(5)Use a similar process to import, clean, and organize the data for Professor Trash Wheel;
(5a)To keep track of which Trash Wheel is which, you may need to add an additional variable to both datasets before combining.
```{r load clean Prof Trash Wheel}
professortrashwheel_df =
  readxl::read_excel(
    "../../datascience/p8105_hw2_kpp2126/Trash Wheel Collection Data.xlsx",
    sheet = "Professor Trash Wheel",
    skip = 1,
    range = cell_cols("A:M")) %>%
  janitor::clean_names() %>%
  drop_na(dumpster) %>%
  mutate(homes_powered = round(homes_powered),
         homes_powered = as.integer(homes_powered),
         year = as.character(year)) %>% #matched this with trashwheel_df, since we are answering for a single year (2020) I kept it coded as.character but could change for future considerations that require panel of years or increments of years which require as.numeric(year)
  add_column(trash_wheel_type = "professor",
             sports_balls = as.integer(NA)) #added as cautionary measure before binding
```


(6)combine this with the Mr. Trash Wheel dataset to produce a single tidy dataset. 
```{r combine}
#janitor::compare_df_cols(trashwheel_df, professortrashwheel_df) used to check data frames matched before binding
combinedtrashwheel_df = 
  bind_rows(trashwheel_df, professortrashwheel_df)
#skimr::skim(combinedtrashwheel_df)
```

*Write a paragraph about these data; you are encouraged to use inline R. Be sure to note the number of observations in the resulting dataset, and give examples of key variables.

The `combinedtrashwheel_df` has `r nrow(combinedtrashwheel_df)` observations and `r ncol(combinedtrashwheel_df)` variables. Of these `r ncol(combinedtrashwheel_df)` variables, there are `r nrow(trashwheel_df)` dumpsters for Mr Trash Wheel, but for Professor Trash Wheel there are `r nrow(professortrashwheel_df)` dumpsters. The Professor Trash Wheel dumpsters do not have values for sports balls, but both data frames collected the same remaining `r ncol(trashwheel_df)-1` variables. Some other variables describe `weight_tons`, measured in tons, and `volume_cubic_yards`, measured in cubic yards and the `date` those values were recorded. The rest of the variables detail the type of trash, such as `sports_balls` and `glass_bottles`, and lastly a measure of homes that could be powered (`homes_powered`) based on the trash collected and processed.

*For available data, what was the total weight of trash collected by Professor Trash Wheel? 
```{r total weight}
total_weight_mrtrashwheel <- combinedtrashwheel_df %>%
  group_by(trash_wheel_type) %>%
  summarise(sum = sum(weight_tons))
```
The total weight of trash collected by Professor Trash Wheel was (baseR) `r `sum(professortrashwheel_df$weight_tons, na.rm =T)` tons, alternatively using the data frame created by method of tidyverse `r total_weight_mrtrashwheel[2, 2]` tons.

*What was the total number of sports balls collected by Mr. Trash Wheel in 2020?
```{r total balls}
total_balls_2020 <- combinedtrashwheel_df %>%
  group_by(trash_wheel_type) %>%
  filter(year == "2020") %>%
  summarise(sum = sum(sports_balls))
```
The total number of sports balls collected by Mr. Trash Wheel in 2020 was `r total_balls_2020[1, 2]` balls.

## Problem 3

This problem uses the FiveThirtyEight data; these data were gathered to create the interactive graphic on this page. In particular, we’ll use the data in pols-month.csv, unemployment.csv, and snp.csv. 

*Our goal is to merge these into a single data frame using year and month as keys across datasets.*

*First, clean the data in pols-month.csv. 
(1)Use `separate()` to break up the variable `mon` into integer variables `year`, `month`, and `day`; 
(2)replace month number with month name using `month.abb`; 
(3)create a `president` variable taking values gop and dem;
(4)remove `prez_dem` and `prez_gop` using `starts_with()`;
(5)remove the `day` variable using `select()`
```{r pols}
pols_df = 
  read_csv("../../datascience/p8105_hw2_kpp2126/pols-month.csv",
           show_col_types = FALSE) %>% #was getting a warning for deliminator for my months so I set it to false
  janitor::clean_names() %>%
  separate(mon, into = c("year", "month", "day"), convert=T) %>%
  mutate(month = month.abb[month], #I used abbreviation to have tidier code so I don't have to recode the unemployment.csv month variables
         president = recode(prez_gop, #curious if set_value_labels would be any different here
                            "0" = "dem", 
                            "1" = "gop", 
                            "2" = "gop")) %>% 
  select(-starts_with("prez"), -day)
```

*Second, clean the data in snp.csv using a similar process to the above. 
(1)arrange according to year and month;
(2)organize so that year and month are the leading columns.
```{r snp}
snp_df =
  read_csv("../../datascience/p8105_hw2_kpp2126/snp.csv",
           show_col_types = FALSE) %>% #was getting a warning for deliminator for my months so I set it to false
  janitor::clean_names() %>%
  separate(date, into = c("year", "month", "day"), convert=T) %>%
  mutate(month = month.abb[month],
         year90s = year + 1900,
         year00s = year + 2000,
         year = case_when(year > 45 ~ year90s, T ~ year00s)) %>%
  arrange(year, month) %>%
  select(year, month, close)
```

*Third, tidy the unemployment data so that it can be merged with the previous datasets. 
(1)This process will involve switching from “wide” to “long” format using `pivot_longer`; 
(2)ensuring that key variables have the same name;
(3)ensuring that key variables take the same values.
```{r unemployment}
unemploy_df =
  read_csv("../../datascience/p8105_hw2_kpp2126/unemployment.csv",
           show_col_types = FALSE) %>% #was getting a warning for deliminator for my months so I set it to false
  rename(year = Year) %>%
  pivot_longer(Jan:Dec,
               names_to = "month", 
               values_to = "unemployment")
```

*Join the datasets by merging snp into pols, and merging unemployment into the result.
```{r left join}
#janitor::compare_df_cols(pols_df, snp_df)
combinedp3_df = 
  left_join(pols_df, snp_df) %>%
  left_join(., unemploy_df)

str(combinedp3_df)
```

*Write a short paragraph about these datasets. 
(1)Explain briefly what each dataset contained;
(2)describe the resulting dataset (e.g. give the dimension, range of years, and names of key variables).
The `pols_df` contains `r nrow(pols_df)` observations and `r ncol(pols_df)` variables that detail the political party (democrat or republican) for `president` (binary), governors `gov_` (continuous), senators `sen_` (continuous), and representatives `rep_` (continuous) between the years 1947-2015. The `snp_df` contains `r nrow(snp_df)` observations and `r ncol(snp_df)` variables. The `close` (continuous) variable indicates the closing value of the Standard & Poor's stock index on the corresponding date between years 2001-2012. The `unemploy_df` has `r nrow(unemploy_df)` observations and `r ncol(unemploy_df)` variables. The `unemployment` variable details the percentage of unemployment on the corresponding year and month, between years 1948-2015.

The resulting data set `combinedp3_df` has `r nrow(combinedp3_df)` observations and `r ncol(combinedp3_df)` variables. The combined data set contains the same variables previously described but now has a combined range of years starting from 1947-2015. I used a left join to ensure that the larger data frame `pols_df` did not lose any observations when joining with the smaller data frame `snp_df`. The `unemployment` and `close` variables were added to the existing `pols_df` variables so that now there is a corresponding row of information related to `president` and other government positions' political parties.

